Thermal reactionomes reveal adaptive responses to thermal

### 2014-10-06 doc/ApTranscriptome_manuscript.Rmd
revisions to ms, mostly Discussion

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage (...refs...). Thermal defense strategies are shaped by the environmental regime organisms experience (...ref...), and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the molecular basis of this variation in thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes. This information is critical for understanding the adaptive potential of species to future climates [@hoffman2008].

@@

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences significantly warmer and less seasonally variable temperatures than *A. picea* (...data...), with correspondingly higher minimal and maximal critical thermal limits (...CTmin and CTmax from HF/DF...). Experimental warming experiments indicate that *Aphaenogaster* abundance and competitive ability may be significantly impacted by temperature increases (...refs...).

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material). We used these data to test two specific predictions. First, we predicted that species from more thermally-stressful environments rely to a greater extent upon constitutive tolerance than plastically-induced resistance mechanisms relative to species from cooler environments. [@barshis2013]. This pattern could be manifested through the number of thermally-responsive genes, or their sensitivity to changes in temperature. Finding that fewer genes are thermally responsive in the warmer-climate species *A. carolinensis* would support the former, while finding that more extreme temperatures are required to induce changes in gene expression in *A. carolinensis* would support the latter. Second, we predicted that long-term adaptive changes to temperatures would involve the same genes that are involved in short-term physiological acclimation (e.g. genetic assimilation [@pigliucci2006]). Finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this prediction.

@@

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally-responsive at a 5% False Discovery Rate [@benjamini1995]. Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach. The result is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate is a lower-bound for the true proportion of genes that are thermally-responsive.

```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally-responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally-responsive in one species but not the other."}

@@

pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
             big.mark = ",", split.tables = Inf, 
             caption = "Table of the number of thermally-responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally-responsive in one species but not the other.")
```

Transcripts varied both qualitatively and quantitatively in their pattern of response to the temperature gradient. We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).

@@

# Shifts in the expression patterns of thermally-responsive transcripts between species

Of the thermally-responsive transcripts, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species. This was not due to a species difference in the overall number of responsive genes (...stat...). Instead, most of these were transcripts that changed from one expression category to another (i.e. the number of transcripts in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1). These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homotranscriptity ...)

@@

## Species differences in the sensitivity of thermally-responsive transcripts to changes in temperature

@@

## Adaptive changes in basal expression are not linked to transcripts involved in short-term physiological acclimation

We tested the hypothesis of adaptation by genetic assimilation, the process by which transcripts change from being environmentally induced to constitutively expressed, in multiple comparisons. First, we predicted that consistent exposure to a temperature stressor should drive a shift from inducibility under stress conditions to continuously elevated expression (e.g. "frontloading", [@barshis2013]). Thus, the warm-adaptated *A. carolinensis* should display higher basal expression of **High** transcripts than *A. picea*, while the reverse should be true for **Low** transcripts.  Instead, we found that transcript expression at optimal temperatures (about 19.5°C) was always greater in *A. carolinensis* than in *A. picea*. This pattern was true for transcripts that had **High** expression in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`), **Low** expression in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`), and **Intermediate** transcripts in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`).

In general, thermally-responsive transcripts in *A. carolinensis* had higher constitutive expression than those in *A. picea*.

The only exception was the set of genes that had **High** or **Bimodal** expression in *A. picea* but were not thermally-responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1). Rather than being constitutively elevated as predicted, these genes were not expressed (...mean TPM...) in *A. carolinensis*. Overall, we found no support for genetic assimilation as the primary mechanism for adaptive shifts in thermal tolerance.

## Candidate genes

We examined the reaction norms for transcripts in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally-responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`). In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A. picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*.

@@

By using a regression-approach, we are able to quantitatively describe changes in transcript expression, rather than simply report the numbers of differentially-expressed transcripts. We found that about 70% of the thermally-responsive transcripts (7,202) have non-linear responses to temperature.

This result is consistent with the only other work we are aware of that examines reaction norms of gene expression, which found that most genes in yeast have have expression differences in response to copper exposure [@hodgins-davis2012]. Finding that thousands of genes responded to temperature highlights the limitations of previous studies [@meyer2011; @barshis2013; @palumbi2014; @oneil2014] that used only a few temperature levels and found that only a few dozen to hundred genes are involved in thermal responses. Underlying the importance of thermal tolerance as a physiological trait [@kingsolver2009], the genetic basis of thermal tolerance is likely to be much more complex than indicated by these previous studies.

The patterns of transcript expression between species differed substantially, with more transcripts down-regulated in *A. carolinensis* and more up-regulated in *A. picea* in response to thermal extremes (Table 1, Fig. 1). The results of gene set enrichment analysis give an indication of what might drive this difference. The transcripts that are shut-down by *A. carolinensis* in response to cold temperatures, but up-regulated by *A. picea* include gene ontology clusters involved in metabolism, transport and regulation. Thus, *A. carolinensis*, the more southerly species, shuts-down metabolism to a greater extent than *A. picea* and so must rely on other mechanisms such as behavior or increased proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we found that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result supports our expectation that *A. picea* would rely to a greater extent upon activation of the cellular stress response and other protective genes to a greater extent than *A. carolinensis*.

Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance. Although the heat-shock response is one of the most conserved across living organisms (ref), it is energetically expensive, particularly under chronic stress conditions (ref).  Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher basal level of chaperone proteins (ref) or by increasing the thermal stability of proteins at the expense of catalytic activity (Arnold et al. 2001).  Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.  Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.

```{r max_exp_PDF, eval = FALSE, echo = FALSE, warning = FALSE, message = FALSE, fig.cap = "Probability density function of the temperature of maximum expression for all thermally-responsive transcripts in each species."}
# reshape data
Ap.df$Species <- ifelse(Ap.df$colony == "ApVT", "A. picea", "A. carolinensis")

maxexplot <- ggplot(Ap.df, aes(x=max.val, fill=Species)) + 
  geom_density(alpha=0.2, position="identity") + 
  labs(x = "Temperature of maximum expression", y = "Density")

print(maxexplot)

####################################
### CHANGE TO REACTION NORM FOR HSPs 
####################################

```

Given the differences in the number of thermally-responsive transcripts between the two species (Fig. 1), it is worth noting the similarity in expression patterns between species. In both species, there are 3-4 $\times$ more transcripts upregulated at cold than high temperatures  (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. At least among *Drosophila* taxa, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. In this light, for *A. picea* and *A. carolinensis*, our results suggest that adaptability to increased temperatures is more limited than to cold temperatures, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].

In contrast to the predictions of the genetic assimilation hypothesis that adaptation occurs via changes from plastically-induced to constitutive expression (e.g. genetic assimilation), we found lower basal transcript expression in *A. picea* for transcripts that were up-regulated in response to cold temperatures (**Low** transcripts) in *A. carolinensis*. We also observed greater basal expression in *A. carolinensis* for **High** transcripts in *A. picea* as well as for transcripts that were **Intermediate** in both species. A potential explanation for the consistently higher basal transcript expression in the warmer-environment species, *A. carolinensis*, is that this species has a higher metabolic rate and thus requires greater transcript expression.

As with all *de novo* transcriptomes, our work has some caveats. First, given that all sequenced ant genomes to date have about only 18,000 genes [@gadau2012], our transcriptome is highly fragmented. Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts. A genome assembly, in progress, will help to reduce fragmentation. Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for distantly-related model organisms *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005]. Large differences in patterns of gene expression may have only small effects on organismal fitness. Given these caveats, transcriptomic studies promise to offer many new insights into how organisms respond and will respond to environmental change, especially for long-lived species that are intractable to study in the lab.

An additional concern is differences in gene expression among species due to drift. While we do in fact expect some genes to differ in expession due to neutral processes, if this was the primary driver of differences among species we would expect to see equal distribution of counts in the off-diagonals of the matched observations (Fig. 1). Instead we find that certain categories are over-represented, and that the differences in the reaction norm shapes are consistent with *a priori* expectations.

@@

In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A. carolinensis*. At least 10% of their transcriptomes are thermally-responsive. For both species, more genes responded to cold than hot temperatures. Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential in response to hot than cold temperatures. Our results also indicate that these two ant species have different responses to thermal extremes. *A. picea* responds by increasing metabolism and activating heat shock proteins and other protective molecules, whereas *A. carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolerance. The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species. Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.

@@

We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina (36.037° N -78.874° W). The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice Bacon DeMarco, pers. communication). Colonies of these species were maintained in the lab for 6 months prior to experimentation.

@@

## Thermally-responsive transcripts

@@

We identified thermally-responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).

>>>>

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Many species, especially small insects, live at temperatures close to their thermal limits and are at risk from current increases in temperature [@deutsch2008; @kingsolver2013; @sunday2014]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, depends on a complex interplay between the structural integrity of cellular components and activiation of physiological response mechanisms to prevent and/or repair damage [@huey1989; @richter2010]. Thermal defense strategies are shaped by the environmental regime organisms experience [@gilchrist1995; @angiletta2003] and thermal limits vary considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. Outside of candidate genes [@krebs1998], little is known about the molecular basis of this variation in thermal tolerance or the link between short-term physiological acclimation and longer-term adaptation to novel temperature regimes. This information is critical for understanding the adaptive potential of species to future climates [@hoffman2008].

@@

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences significantly warmer and less seasonally variable temperatures than *A. picea* (...data...), with correspondingly higher minimal and maximal critical thermal limits (...CTmin and CTmax from HF/DF...). Experimental warming experiments indicate that *Aphaenogaster* abundance and competitive ability may be significantly impacted by temperature increases (...refs...).

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material). We used these data to test two specific predictions. First, we predicted that species from more thermally-stressful environments rely to a greater extent upon constitutive tolerance (e.g. "frontloading", [@barshis2013]) than plastically-induced resistance mechanisms relative to species from cooler environments. This pattern could be manifested through the number of thermally responsive genes, or their sensitivity to changes in temperature. Finding that fewer genes are thermally responsive in the warmer-climate species *A. carolinensis* would support the former, while finding that more extreme temperatures are required to induce changes in gene expression in *A. carolinensis* would support the latter. Second, we predicted that long-term adaptive changes to temperatures would involve the same genes that are involved in short-term physiological acclimation (e.g. genetic assimilation [@pigliucci2006]). Specifically, finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased constitutive expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this prediction.

@@

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995]. Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach. The result is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate is a lower-bound for the true proportion of genes that are thermally responsive.

```{r type_table, echo=FALSE, results='asis', fig.cap = "Table of the number of thermally responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally responsive in one species but not the other."}

@@

pandoc.table(Ap.type.table[,c("Low", "Intermediate", "High", "Bimodal", "NotResp")], style="rmarkdown",
             big.mark = ",", split.tables = Inf, 
             caption = "Table of the number of thermally responsive transcripts by expression type for *A. carolinensis* and *A. picea*. **Low** are transcripts with increased expression at low temperatuers (< 10°C), **Intermediate** are transcripts with maximum expression between 10 - 30°C, **High** are transcripts with increased expression at high temperatures (> 31°C), **Bimodal** are transcripts with increased expression at both low and high temperatures, while **NotResp** are transcripts that are thermally responsive in one species but not the other.")
```

Transcripts varied both qualitatively and quantitatively in their pattern of response to the temperature gradient. We grouped the thermally responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C).

@@

# Shifts in the expression patterns of thermally responsive transcripts between species

Of the thermally responsive transcripts, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) had an expression pattern that differed between the two species. This was not due to a species difference in the overall number of responsive genes (...stat...). Instead, most of these were transcripts that changed from one expression category to another (i.e. the number of transcripts in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Table 1). These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homotranscriptity ...) indicating that these differences are not simply due to stochastic processes.

@@

## Species differences in the sensitivity of thermally responsive transcripts to changes in temperature

@@

## Adaptive changes in constitutive expression are not linked to transcripts involved in short-term physiological acclimation

We tested the hypothesis of adaptation by genetic assimilation, the process by which transcripts change from being environmentally induced to constitutively expressed, in multiple comparisons. First, we predicted that consistent exposure to a temperature stressor should drive a shift from inducibility under stress conditions to continuously elevated expression. Thus, the warm-adaptated *A. carolinensis* should display higher constitutive expression of **High** transcripts than *A. picea*, while the reverse should be true for **Low** transcripts.  Instead, we found that transcript expression at optimal temperatures (about 19.5°C) was always greater in *A. carolinensis* than in *A. picea*. This pattern was true for transcripts that had **High** expression in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`), **Low** expression in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`), and **Intermediate** transcripts in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`).

In general, thermally responsive transcripts in *A. carolinensis* had higher constitutive expression than those in *A. picea*.

The only exception was the set of genes that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1). Rather than being constitutively elevated as predicted, these genes were not expressed (...mean TPM...) in *A. carolinensis*. Overall, we found no support for genetic assimilation as the primary mechanism for adaptive shifts in thermal tolerance.

## Greater up-regulation of heat shock proteins in the cool-climate species *A. picea*

We examined the reaction norms for transcripts in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`). In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A. picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*.

@@

Species persistence in face of climate change requires both short-term physiological acclimation and longer-term adapative changes. In this study, we explored the relationship between these processes using whole organism gene expression data from closely-related ant species.

Recent RNA-seq studies of thermal tolerance, using only a few temperature conditions, have identified a few dozen to hundred genes that are differentially expressed between conditions [@meyer2011; @barshis2013; @palumbi2014; @oneil2014]. By applying a regression approach, we were able to quantitatively describe changes in transcript expression across a thermal gradient. In particular, by characterizing the thermal reactionomes of the ant species *Aphaenogaster picea* and *A. carolinensis* we found differences indicative of adaptation to climate, even while the total number of differentially-expressed genes was similar between the two species (Table 1). In total, we found that about 8% of transcripts (~9,000) are thermally responsive in each species, and that about 70% of these transcripts have a non-linear response to temperature that differs between species, which would not have been detected by standard differential expression approaches. Generally, this indicates that the molecular response to thermal extremes is much more complex than indicated by previous studies, as would be expected given the ecological importance of thermal tolerance [@kingsolver2009].

Following the genetic assimilation hypothesis that adaptation occurs via changes from plastically induced to constitutive expression, we predicted a tradeoff between constitutive expression levels and inducibility of a transcript. In contrast to this prediction, constitutive expression levels at 19.5°C were higher in *A. carolinensis* than *A. picea* in nearly all comparisons. The exception to this was for genes that had **High** or **Bimodal** expression in *A. picea* and were not responsive in *A. carolinensis*. Rather than being constitutively elevated as predicted, these genes were not expressed (...mean TPM...) in *A. carolinensis*. ...gene ontology terms...genes recruited to thermal tolerance... Consistent with our results but at a much shorter time-scale, no evidence for genetic assimilation at the expression level was found in an experimental evolution study with *Caenorhabditis remanei* selected for 10 generations to increased heat tolerance, even though there was a substantial increase in heat tolerance and loss of plasticity in the evolved populations [@sikkink2014]. In this study, the genetic assimilation at the phenotypic level was caused by a shift in the threshold for temperature response, similar to what we observed here (Fig. 2).

While our results did not the genetic assimilation hypothesis, the observation of higher constitutive expression of thermally responsive transcripts is consistent with the prediction that the warm-climate species *A. carolinensis* would rely more upon "frontloading" than the cool-climate *A. picea*. Moreover, *A. carolinensis* shut-down expression of more transcripts (Fig. 2) at thermal extremes, and had greater inertia of expression to changes in temperature (Fig. 2) than did *A. picea*. In constrast, the cool-climate species *A. picea* activated more transcripts in response to both thermal extremes, including enrichment of biological processes related to metabolism, transport and regulation. One possible explanation for this difference in responses is that *A. picea* mounts a classical cellular stress response [@kultz2005; @richter2010], requiring increased metabolism, while *A. carolinensis* relies upon constitutive defenses [@barshis2013], proteome stability [@ghosh2010] or behavior [@sunday2014] to tolerate thermal stress. Consistent with this hypothesis, we found that of the 7 thermally responsive heat shock proteins, all 7 increased expression in response to thermal extremes in *A. picea*, but only 3 do in *A. carolinensis*. In contrast, by shutting-down molecular activity, must rely upon other mechanisms such as behavior or increased proteome stability to tolerate thermal stress. These differences are in line with the expectations if *A. picea*, a northern species with fewer good foraging days, must forage in suboptimal conditions whereas *A. carolinensis* can, with a longer season, afford to restrain from foraging in suboptimal conditions. In fact, restrained foraging under stressful conditions by the red harvester ant *Pogonomyrmex barbatus* has been shown to increase colony fitness [@gordon2013].

Given the differences in the patterns thermal-responsiveness between species (Fig. 2), it is worth noting the similarity in expression patterns between species. In both species, there are 3-4 $\times$ more transcripts upregulated at cold than high temperatures  (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. At least among *Drosophila* taxa, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. In this light, for *A. picea* and *A. carolinensis*, our results suggest that adaptability to increased temperatures is more limited than to cold temperatures, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].

A major contribution of this study is the reaction norm approach for gene expression data (but see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach using differential expression among many levels). A number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed genes are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], identify genes that are correlated, but not necessarily causative, with phenotypes [@feder2005] and are prone to "storytelling" [@pavlidis2013]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive genes in each species. Thus, while we use gene ontology information to explain our main results, the key findings about the number of responsive genes and shifts in temperatures of induction do not depend on functional annotation. Similarly, in the plant and animal breeding literature there has been a shift from using genome-wide association methods which are prone to the same problems listed above, to using genomic selection methods which provide a direct means to predict a population's response to selection [@goddard2007].

As with all *de novo* transcriptomes, our work has some caveats. First, given that all sequenced ant genomes to date have about only 18,000 genes [@gadau2012], our transcriptome is highly fragmented. Although we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts. A genome assembly, in progress, will help to reduce fragmentation. Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is not as good as it would be for distantly-related model organisms *Drosophila* and *Apis*. This is a problem that plagues  Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005]. Large differences in patterns of gene expression may have only small effects on organismal fitness.

@@

Our results are congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance. Although the heat-shock response is one of the most conserved across living organisms [@kultz2005], it is energetically expensive, particularly under chronic stress conditions (.....) .  Under such circumstances, it may be advantageous to proactively prevent thermal damage even at the cost of reduced metabolic efficiency, either by maintaining a higher constitutive level of chaperone proteins (ref) or by increasing the thermal stability of proteins at the expense of catalytic activity (Arnold et al. 2001).  Thus, although in the short term increasing temperature stress leads to a quantitatively stronger induced response, adapting to such stress over evolutionary time appears to require a qualitative shift in mechanism of resistance that can alter not only the magnitude, but the sign of gene expression change in response to temperature.  Whether such a shift would be possible in the compressed time frame of projected climate change, particularly for long-lived organisms such as ants, is likely to be critical in determining the capacity of populations to adapt to more frequent and long-lasting stressors.

In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A. carolinensis*. At least 10% of their transcriptomes are thermally responsive. For both species, more genes responded to cold than hot temperatures. Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential in response to hot than cold temperatures. Our results also indicate that these two ant species have different responses to thermal extremes. *A. picea* responds by increasing metabolism and activating heat shock proteins and other protective molecules, whereas *A. carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolerance. The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species. Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.

@@

We used colonies of the ant species *Aphaenogaster picea*, collected from Molly Bog, Vermont (University of Vermont Natural Areas; 44.508°N, -72.702°W) and *Aphaenogaster carolinensis*, collected from Durham, North Carolina (36.037° N -78.874° W). The phylogeny of *Aphaeogaster* is currently unresolved [@umphrey1996], but putative species identification was possible based on morphology (Bernice DeMarco, pers. communication), and there is about 2% pairwise divergence between these clades using five nuclear genes (Bernice DeMarco, unpublished data). Colonies of these species were maintained in the lab for 6 months prior to experimentation.

@@

## Thermally responsive transcripts

@@

We identified thermally responsive transcripts as the subset under the FDR threshold that had significant $\beta_2(temp)$, $\beta_3(temp^2)$, $\beta_4(species * temp)$ or $\beta_5(species * temp^2)$ terms after step-wise model selection by AIC. All analyses were performed in R [@r_core_team2013] and are fully reproducible (Supplemental file).
